Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-07-27 threshold/draft.Rnw
Better references

@@

Efficiently computing a subset of a correlation matrix consisting of values
above a specified threshold is important to many practical applications.
Real-world problems in genomics, machine learning, finance, and other fields
can produce correlation matrices too large to explicitly form and tracticably
compute. Often, only values corresponding to highly-correlated vectors are of
interest, and those values typically make up a small fraction of the overall
correlation matrix. We present a method based on the singular value
decomposition (SVD) and its relationship to the data covariance structure that
can efficiently compute thresholded subsets of very large correlation matrices.

@@

XXX add discussion of applications (genomics?) and previous work (especially
\cite{xstk1},\cite{xstk2},\cite{xbm} and
related but different work that use PCA: \cite{latj} and \cite{bhpt} are related but different),
then on to the preliminaries:
\[\]

@@

In practice, many pairs of vectors may be ruled out in this fashion by
inspecting distances in low-dimensional subspaces, substantially reducing the
complexity of computing thresholded correlation matrices.  Approaches that
cheaply omit pairs of vectors from consideration are called \emph{pruning
methods}, and a variety of pruning methods have been described elsewhere for
the thresholded correlation problem; see, for
example\cite{mnl,xbm,xstk1,xstk2,zf} and the references therein.  The approach
described here prunes along an intuitive path of decreasing variance defined by
the data and computed by the SVD.

@@

Pairs of vectors below the correlation threshold that lie more than one index
difference apart relative to the permutation $P$ can be similarly pruned by
replacing $D^1$ with $D^j$, $j=1,2,\ldots,\ell$ in the above expressions.
Following this process, we can produce a well-pruned candidate set of possible
vector pairs that meet a given correlation threshold using $\ell$ matrix vector
products with matrices of order $n \times p$, where $p$ is chosen in practice
such that $p<<k$. Setting $p=10$ for the example shown in Figure
\ref{fig1} prunes all but $431$ possible vector pairs out of a total
$419(419-1)/2 = 87,571$ with $\ell=138$.

@@

The augmented implicitly restarted Lanczos bidiagonalization algorithm (IRLBA)
of Baglama and Reichel \cite{br} efficiently computes truncated singular value
decompositions of large dense or sparse matrices. Reference implementations are
available for R\cite{irlba-R}, Matlab\cite{irlba-M}, and Python\cite{irlba-P}.
The R implementation includes easy to use and efficient methods for restarting
to increase the subspace dimension (see step 5 of Algorithm \ref{prune}), and
for computing the SVD of scaled and centered matrices without explicitly
forming a new matrix.

We relax the column mean and scale assumptions without introducing much additional
computational or storage overhead as follows.  Let
$z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
of the matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$ be the vector
of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$ a diagonal
matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
matrix with zero column means, and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
\]
The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
products. The main idea behind efficient application of IRLB to correlation
problems replaces matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ in
the iterations, implicitly working with a scaled and centered matrix without
forming it.  This comes at the cost of storing two additional length $n$ vectors and
computing two additional vector inner products per matrix-vector product.

@@

%% XXX switch to bibtex / make citations uniform

\begin{thebibliography}{99}
\bibitem{br}
Baglama, J. and Reichel, L., Augmented Implicitly Restarted Lanczos Bidiagonalization Methods, SIAM J. Sci. Comput., 2005.
\bibitem{bhpt}
Bair, E., Hastie, T., Paul, D.., and Tibshirani, P., Prediction by supervised principal components, Journal of the American Statistical Association 101, no. 473, 2006.
\bibitem{latj}
Levin, Asriel U., Todd K. Leen, and John E. Moody. Fast pruning using principal components. (1993).
\bibitem{mnl}
Mueen, A., Nath, S., and Liu, J., Fast approximate correlation for
massive time-series data, in Proceedings of the 2010 ACM SIGMOD International
Conference on Management of data, 2010, pp. 171-182.
\bibitem{xbm}
Xiong, Hui, Mark Brodie, and Sheng Ma. "Top-cop: Mining top-k strongly correlated pairs in large databases." Data Mining, 2006. ICDM'06. Sixth International Conference on. IEEE, 2006.
\bibitem{xstk1}
Xiong, Hui, et al. "Exploiting a support-based upper bound of Pearson's correlation coefficient for efficiently identifying strongly correlated pairs." Proceedings of the tenth ACM SIGKDD international conference on Knowledge discovery and data mining. ACM, 2004.
\bibitem{xstk2}
Xiong, Hui, et al. "TAPER: A two-step approach for all-strong-pairs correlation query in large databases." Knowledge and Data Engineering, IEEE Transactions on 18.4 (2006): 493-508.
\bibitem{zf}
Zhang, J. and Feigenbaum, J, Finding Highly Correlated Pairs Efficiently with Powerful Pruning,
in Proceedings of ACM Conference on Information and Knowledge Management (CIKM), 2006, pp. 152-161.
\bibitem{irlba-R}
\url{https://cran.r-project.org/web/packages/irlba/index.html}
\bibitem{irlba-P}
\url{https://github.com/bwlewis/irlbpy}
\bibitem{irlba-M}
\url{http://www.math.uri.edu/~jbaglama/software/irlba.m}
\end{thebibliography}
\end{document}

>>>>

%%<<eval=TRUE, echo=FALSE>>=
%%Sys.setenv(TEXINPUTS=getwd(),
%%           BIBINPUTS=getwd(),
%%           BSTINPUTS=getwd())
%%@

@@

Efficiently computing a subset of a correlation matrix consisting of values
above a specified threshold is important to many practical applications.
Real-world problems in genomics, machine learning, finance, and other fields
can produce correlation matrices too large to explicitly form and tractably
compute. Often, only values corresponding to highly-correlated vectors are of
interest, and those values typically make up a small fraction of the overall
correlation matrix. We present a method based on the singular value
decomposition (SVD) and its relationship to the data covariance structure that
can efficiently compute thresholded subsets of very large correlation matrices.

@@

XXX
\\

@@

In practice, many pairs of vectors may be ruled out in this fashion by
inspecting distances in low-dimensional subspaces, substantially reducing the
complexity of computing thresholded correlation matrices.  Approaches that
cheaply omit pairs of vectors from consideration are called \emph{pruning
methods}, and a variety of pruning methods have been described elsewhere for
the thresholded correlation problem; see, for
example\cite{prune1, prune2, prune3, prune4}
and the references therein.  The approach described here prunes along an
intuitive path of decreasing variance defined by the data and computed by the
SVD.

@@

Pairs of vectors below the correlation threshold that lie more than one index
difference apart relative to the permutation $P$ can be similarly pruned by
replacing $D^1$ with $D^j$, $j=2,3,\ldots,\ell$ in the above expressions.
Following this process, we can produce a well-pruned candidate set of possible
vector pairs that meet a given correlation threshold using $\ell$ matrix vector
products with matrices of order $n \times p$, where $p$ is chosen in practice
such that $p<<k$. Setting $p=10$ for the example shown in Figure
\ref{fig1} prunes all but $431$ possible vector pairs out of a total
$419(419-1)/2 = 87,571$ with $\ell=138$.

@@

The augmented implicitly restarted Lanczos bidiagonalization algorithm (IRLBA)
of Baglama and Reichel \cite{irlba} efficiently computes truncated singular value
decompositions of large dense or sparse matrices. Reference implementations are
available for R\cite{irlbar}, Matlab\cite{irlbam}, and Python\cite{irlbap}.

We relax the column mean and scale assumptions without introducing much additional
computational or storage overhead as follows.  Let
$z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
of the matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$ be the vector
of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$ a diagonal
matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
matrix with zero column means and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
\]
The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
products. The main idea behind efficient application of IRLB to correlation
problems replaces matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ in
the iterations, implicitly working with a scaled and centered matrix without
forming it.  This comes at the cost of storing two additional length $n$ vectors and
at most computing two additional vector inner products per matrix-vector product.

@@

\begin{thebibliography}{99}

%% Related use of SVD in other methods
\bibitem{spca}
Bair, Eric, et al. Prediction by supervised principal components. Journal of the American Statistical Association 101.473 (2006).

%% irlba
\bibitem{irlba}
Baglama, James, and Lothar Reichel. Augmented implicitly restarted Lanczos bidiagonalization methods. SIAM Journal on Scientific Computing 27.1 (2005): 19-42.
\bibitem{irlbam}
Baglama, James,  {\url{http://www.math.uri.edu/~jbaglama/software/irlba.m}}.

%% Applications in machine learning
\bibitem{ml1}
Hall, Mark A. Correlation-based feature selection for machine learning. Diss. The University of Waikato, 1999.

%% IRLBA implementations
\bibitem{irlbap}
Kane, Michael and Lewis, Bryan,  \url{https://github.com/bwlewis/irlbpy}.

%% Related use of SVD in neural networks
\bibitem{neural}
Levin, Asriel U., Todd K. Leen, and John E. Moody. Fast pruning using principal components. (1993).

%% IRLBA implementations
\bibitem{irlbar}
Lewis, Bryan,  \url{https://cran.r-project.org/web/packages/irlba}.

%% Approximate correlation
\bibitem{approxcor}
Mueen, Abdullah, Suman Nath, and Jie Liu. Fast approximate correlation for massive time-series data. Proceedings of the 2010 ACM SIGMOD International Conference on Management of data. ACM, 2010.

%% Applications in genomics
\bibitem{genomics}
Song, Lin, Peter Langfelder, and Steve Horvath. Comparison of co-expression measures: mutual information, correlation, and model based indices. BMC bioinformatics 13.1 (2012): 328.

%% Other correlation pruning methods
\bibitem{prune1}
Xiong, Hui, et al. TAPER: A two-step approach for all-strong-pairs correlation query in large databases. Knowledge and Data Engineering, IEEE Transactions on 18.4 (2006): 493-508.
\bibitem{prune2}
Xiong, Hui, Mark Brodie, and Sheng Ma. Top-cop: Mining top-k strongly correlated pairs in large databases. Data Mining, 2006. ICDM'06. Sixth International Conference on. IEEE, 2006.

%% Applications in machine learning
\bibitem{ml2}
Yu, Lei, and Huan Liu. Feature selection for high-dimensional data: A fast correlation-based filter solution. ICML. Vol. 3. 2003.

%% Other correlation pruning methods
\bibitem{prune3}
Zhang, Jian, and Joan Feigenbaum. Finding highly correlated pairs efficiently with powerful pruning. Proceedings of the 15th ACM international conference on Information and knowledge management. ACM, 2006.
\bibitem{prune4}
Zhu, Shiwei, et al. Scaling up top-K cosine similarity search. Data \& Knowledge Engineering 70.1 (2011): 60-83.

\end{thebibliography}
\end{document}
